Endogenous Politics and \\ the Design of Trade Institutions

### 2018-03-02 presentation.tex
final presentation (handout) for UNL

\pause
\begin{enumerate}[<+->]
\item Can trade agreements be used to manipulate domestic lobbying incentives?
	\begin{itemize}
		\item Government objective function
	\end{itemize}
\vskip.1in
	\item When is endogenizing political pressure important for answering optimal design questions?
	\begin{itemize}
		\item Exogenous vs. endogenous politics
	\end{itemize}
\end{enumerate}

@@

\pause
\begin{itemize}[<+->]
	\item TAs may be used to manipulate domestic political actors (no long-run distortions)
	\item Simple modeling framework can capture results from models of both exogenous and endogenous politics
		\begin{itemize}
			\item For both tariff caps and escape clauses, endogenous politics changes outcomes dramatically
		\end{itemize}
	\item Standard, theoretical escape clause can't work in the presence of endogenous political pressure
		\begin{itemize}
			\item Points to real-world design of WTO Agreement on Safeguards
			\item May explain why escape clause has fallen out of use
		\end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}{Applied Tariff Decision}
\pause
  Baldwin-style government objective function:
\pause
\[
  W = \mathit{CS}_X(\tau) + \ga(s,e) \pi_X(\tau) + \mathit{CS}_Y(\tau^*) + \pi_Y(\tau^*) + \mathit{TR}(\tau)
\]
\vskip-.1in
\pause
\begin{itemize}[<+->]
	\item Standard \textit{except} weight on import-competing industry profits $\ga(s,e)$: 
		\begin{itemize}
			\item $s$: exogenous shock
			\item $e$: lobbying effort
		\end{itemize}
	\item Optimal applied tariff is a function of $\ga(s,e)$
		\begin{itemize}
			\item Ignores foreign welfare
			\item Takes into account trade agreement enforcement
		\end{itemize}
	\item Assume $\ga$, $\ga^*$ is private info of each government
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Domestic Political Pressure}
Two potential sources
\pause
\begin{enumerate}[<+->]
	\item Exogenous shocks
		\begin{itemize}[<+->]
			\item Shock directly to $\ga$ as in Bagwell $\&$ Staiger (2005): $\ga$, $\ga^*$ with CDF $H(\ga)$ on support $\left[\un{\ga},\ov{\ga}\right]$; or
			\item Can take $\ga$ as a function of $s$: $\ga(s)$
		\end{itemize}

@@

	\item If weights must sum to 1, welfare no longer monotonic in $\ga$
\end{itemize}
\end{frame}

@@

{\begin{frame}{Tariff Caps: Exogenous vs. Endogenous $\ga$}
\pause
Must set tariff at or below specified level (aka tariff cap)
\pause
\begin{itemize}[<+->]
	\item $\ga(s)$, i.e. exogenous: Negotiated weak bindings (a) are higher than those gov'ts would choose if they instead negotiated strong bindings and (b) imply that governments with low realizations of $\ga$ set their applied tariffs strictly below the bound level.
		\begin{itemize}
			\item Bagwell $\&$ Staiger 2005 result
		\end{itemize}

@@

\begin{frame}{When the world is more complicated...}
Now suppose political pressure is a result of both endogenous and exogenous forces (i.e. $\ga(s,e)$):
\pause
\vskip.1in
\begin{beamerboxesrounded}[upper=palette tertiary, shadow=true]{Water in the Bindings with $\ga(s,e)$}
  Assume $\ga(s,e) = \ga(s) + \ga(e)$. In order for governments to set applied tariffs strictly below the weak binding, the low shock must be realized \textit{and} the lobby must not have the incentive to `top up' political pressure to the level associated with the binding. 
\end{beamerboxesrounded}

\pause
\begin{itemize}[<+->]
	\item Can happen if gov't mis-judges lobby's incentives
	\item In general, gov't prefers cap because lobby will `fill in' for low shock up to gov'ts optimal level of $\ga$
\end{itemize}
\end{frame}

@@

\begin{frame}{An Escape Clause for Endogenous Politics}
\pause
Assume a WTO-like set up: gov't can choose between $\tau^a$, `escape' tariff $\tau(s)$, or politically-optimal $\tau$ matched to $\ga(s,e)$
\pause
\begin{itemize}[<+->]
	\item Assume $s$ verifiable, so no punishment for $\tau(s)$
	\item Punishment for $\tau(\ga(s,e)) > \tau(s)$
\end{itemize}

@@

\section{Conclusion}

\begin{comment}
\begin{frame}{Ex-Ante Lobbying}
Without ex-ante: gov't anticipates ex-post lobbying, chooses $\tau^{\text{TA}}$ that maximizes objective function given $e^{\text{EP}}$
  \[
	  \max_\tau W(\ga(e^{\text{EP}}),\tau) + W^*(\tau)
	\]
\pause
When ex-ante lobbying ($e^{\text{EA}}$) is possible, we could have
\pause
  \[
	  \max_\tau W(\ga(e^{\text{EA}}),\tau) + W^*(\tau)
	\]
	but the lobby would now have to pay \textit{twice}.
\end{frame}
\end{comment}

>>>>

\pause
\begin{enumerate}[<+->]
\item Can trade agreements (TAs) be used to manipulate domestic lobbying incentives?
	\begin{itemize}
		\item Government objective function
	\end{itemize}
\vskip.1in
	\item What is the optimal design of various trade agreement properties?
	\begin{itemize}
		\item Exogenous vs. endogenous politics
	\end{itemize}
\end{enumerate}

@@

\pause
\begin{itemize}[<+->]
	\item TAs may be used to manipulate domestic political actors (even with no long-run distortions)
	\item Simple modeling framework can capture results from models of both exogenous and endogenous politics
		\begin{itemize}
			\item For both tariff caps and escape clauses, endogenous politics changes outcomes dramatically
		\end{itemize}
	\item Standard, theoretical escape clause can't work in the presence of endogenous political pressure
		\begin{itemize}
			\item Points to real-world design of WTO Agreement on Safeguards
			\item May explain why escape clause has fallen out of use
		\end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}{Applied Tariff Decision}
\pause
  Baldwin-style government objective function:
\pause
\[
  W = \mathit{CS}_X(\tau) + \ga(s,e) \pi_X(\tau) + \mathit{CS}_Y(\tau^*) + \pi_Y(\tau^*) + \mathit{TR}(\tau)
\]
\vskip-.1in
\pause
\vskip.1in
\begin{itemize}[<+->]
	\item Standard \textit{except} weight on import-competing profits: 
		\begin{itemize}
			\item $s$: exogenous shock
			\item $e$: lobbying effort
		\end{itemize}
	\item Optimal applied tariff is a function of $\ga(s,e)$
		\begin{itemize}
			\item Ignores foreign welfare
			\item Takes into account trade agreement enforcement
		\end{itemize}
	\item Assume $\ga$, $\ga^*$ is private info of each government
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Domestic Political Pressure: Two potential sources}

\pause
\begin{enumerate}[<+->]
	\item Exogenous shocks
		\begin{itemize}[<+->]
			\item Shock directly to $\ga$ as in Bagwell $\&$ Staiger (2005): $\ga$, $\ga^*$ with CDF $H(\ga)$ on support $\left[\un{\ga},\ov{\ga}\right]$; or
			\item Can take $\ga$ as a function of shock $s$: $\ga(s)$
		\end{itemize}

@@

	\item If lobbying effort subtracted as cost from $W$, welfare no longer monotonic in $\ga$
	\pause
		\begin{itemize}
			\item If weights must sum to 1, welfare also not monotonic in $\ga$
		\end{itemize}
\end{itemize}
\end{frame}

@@

{\begin{frame}{Tariff Caps: Exogenous vs. Endogenous $\ga$}
\pause
Must set tariff at or below specified level (aka weak binding)
\pause
\begin{itemize}[<+->]
	\item $\ga(s)$, i.e. exogenous: Negotiated weak bindings (a) are higher than those gov'ts would choose if they instead negotiated strong bindings and (b) imply that governments with low realizations of $\ga$ set their applied tariffs strictly below the bound level.
		\begin{itemize}
			\item Bagwell $\&$ Staiger 2005 result
		\end{itemize}

@@

\begin{frame}{When the world is more complicated...}
Now suppose $\ga(s,e)$, i.e. political pressure is a result of both endogenous and exogenous forces:
\pause
\vskip.1in
\begin{beamerboxesrounded}[upper=palette tertiary, shadow=true]{Water in the Bindings with $\ga(s,e)$}
  Assume $\ga(s,e) = \ga(s) + \ga(e)$. In order for governments to set applied tariffs strictly below the weak binding, the low shock must be realized \textit{and} the lobby must not have the incentive to `top up' political pressure to the level associated with the binding. 
\end{beamerboxesrounded}

\pause
\begin{itemize}[<+->]
	\item Can happen if gov't mis-judges lobby's incentives
	\item In general, gov't prefers cap because lobby will `fill in' for low shock up to gov's optimal level of $\ga$
\end{itemize}
\end{frame}

@@

\begin{frame}{An Escape Clause for a Complicated World}
\pause
Assume a WTO-like set up: gov't can choose between $\tau^a$, `escape' tariff $\tau(s)$, or politically-optimal $\tau$ matched to $\ga(s,e)$
\pause
\begin{itemize}[<+->]
	\item Assume $s$ verifiable, so no punishment for $\tau(s)$
	\item Punishment for $\tau(\ga(s,e)) > \tau(s)$
\end{itemize}

@@

\section{Conclusion}
\subsection{}
\begin{comment}
\begin{frame}{Ex-Ante Lobbying}
Without ex-ante: gov't anticipates ex-post lobbying, chooses $\tau^{\text{TA}}$ that maximizes objective function given $e^{\text{EP}}$
  \[
	  \max_\tau W(\ga(e^{\text{EP}}),\tau) + W^*(\tau)
	\]
\pause
When ex-ante lobbying ($e^{\text{EA}}$) is possible, we could have
\pause
  \[
	  \max_\tau W(\ga(e^{\text{EA}}),\tau) + W^*(\tau)
	\]
	but the lobby would now have to pay \textit{twice}.
\end{frame}
\end{comment}
